---
title: "DATA_318_Final_Project"
author: "Riley Fiske"
date: "4/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(lubridate)
library(caret)
library(leaps) # Needed for Subset Selection
library(ipred) # for bagging
library(rpart) # for tree-based methods
library(gbm) # for boosting
library(ggfortify) # for plotting PCA
library(factoextra) # for plotting PCA
library(pls) # for PCR
library(gridExtra)
library(ggplot2)

citation('tidyverse')
citation('lubridate')
citation('caret')
citation('leaps')
citation('ipred')
citation('rpart')
citation('gbm')
citation('ggfortify')
citation('factoextra')
citation('pls')
citation('gridExtra')
citation('ggplot2')
```

## Project Description
Using Foreign Exchange rates relative to the US Dollar, this project sought to create a model that would most accurately predict the exchange rate of the Euro to the Dollar using lead variables from one and two previous days only. It started with all currencies, but it became obvious quite early that certain currencies were too closely related to the Euro and leakage was occurring. All European currencies were then filtered out, quickly followed by the Australian and New Zealand Dollars since they are also very closely related.

## Data Input and Cleaning
```{r}
Forex_DF <- read.csv("https://raw.githubusercontent.com/rdfiske17/DATA-318/main/Datasets/eurofxref-hist.csv",stringsAsFactors=TRUE)
#https://data.humdata.org/dataset/ecb-fx-rates?force_layout=desktop
#List of Currency Names: https://www.ecb.europa.eu/stats/policy_and_exchange_rates/euro_reference_exchange_rates/html/index.en.html
#UPDATED: use https://www.ecb.europa.eu/stats/eurofxref/eurofxref-hist.zip to download file manually as it cannot be dynamically pulled anymore

Forex_DF <- Forex_DF[-1,] %>% dplyr::select(-TRL) #TRL became TRY in 2005
Forex_DF$Date <- as.character(Forex_DF$Date)
Forex_DF$USD <- as.numeric(as.character(Forex_DF$USD))
Forex_DF$JPY <- as.numeric(as.character(Forex_DF$JPY))
Forex_DF$BGN <- as.numeric(as.character(Forex_DF$BGN))
Forex_DF$CYP <- as.numeric(as.character(Forex_DF$CYP))
Forex_DF$CZK <- as.numeric(as.character(Forex_DF$CZK))
Forex_DF$DKK <- as.numeric(as.character(Forex_DF$DKK))
Forex_DF$EEK <- as.numeric(as.character(Forex_DF$EEK))
Forex_DF$GBP <- as.numeric(as.character(Forex_DF$GBP))
Forex_DF$HUF <- as.numeric(as.character(Forex_DF$HUF))
Forex_DF$LTL <- as.numeric(as.character(Forex_DF$LTL))
Forex_DF$LVL <- as.numeric(as.character(Forex_DF$LVL))
Forex_DF$MTL <- as.numeric(as.character(Forex_DF$MTL))
Forex_DF$PLN <- as.numeric(as.character(Forex_DF$PLN))
Forex_DF$RON <- as.numeric(as.character(Forex_DF$RON))
Forex_DF$ROL <- as.numeric(as.character(Forex_DF$ROL))
Forex_DF$SEK <- as.numeric(as.character(Forex_DF$SEK))
Forex_DF$SIT <- as.numeric(as.character(Forex_DF$SIT))
Forex_DF$SKK <- as.numeric(as.character(Forex_DF$SKK))
Forex_DF$CHF <- as.numeric(as.character(Forex_DF$CHF))
Forex_DF$ISK <- as.numeric(as.character(Forex_DF$ISK))
Forex_DF$NOK <- as.numeric(as.character(Forex_DF$NOK))
Forex_DF$HRK <- as.numeric(as.character(Forex_DF$HRK))
Forex_DF$RUB <- as.numeric(as.character(Forex_DF$RUB))
Forex_DF$TRY <- as.numeric(as.character(Forex_DF$TRY))
Forex_DF$AUD <- as.numeric(as.character(Forex_DF$AUD))
Forex_DF$BRL <- as.numeric(as.character(Forex_DF$BRL))
Forex_DF$CAD <- as.numeric(as.character(Forex_DF$CAD))
Forex_DF$CNY <- as.numeric(as.character(Forex_DF$CNY))
Forex_DF$HKD <- as.numeric(as.character(Forex_DF$HKD))
Forex_DF$IDR <- as.numeric(as.character(Forex_DF$IDR))
Forex_DF$ILS <- as.numeric(as.character(Forex_DF$ILS))
Forex_DF$INR <- as.numeric(as.character(Forex_DF$INR))
Forex_DF$KRW <- as.numeric(as.character(Forex_DF$KRW))
Forex_DF$MXN <- as.numeric(as.character(Forex_DF$MXN))
Forex_DF$MYR <- as.numeric(as.character(Forex_DF$MYR))
Forex_DF$NZD <- as.numeric(as.character(Forex_DF$NZD))
Forex_DF$PHP <- as.numeric(as.character(Forex_DF$PHP))
Forex_DF$SGD <- as.numeric(as.character(Forex_DF$SGD))
Forex_DF$THB <- as.numeric(as.character(Forex_DF$THB))
Forex_DF$ZAR <- as.numeric(as.character(Forex_DF$ZAR))
Forex_DF$Date <- as.Date(Forex_DF$Date)

Forex_DF <- Forex_DF %>%
  mutate(EUR = 1/USD,
         JPY = (JPY*EUR)/USD,
         BGN = (BGN*EUR)/USD,
         CYP = (CYP*EUR)/USD,
         CZK = (CZK*EUR)/USD,
         DKK = (DKK*EUR)/USD,
         EEK = (EEK*EUR)/USD,
         GBP = (GBP*EUR)/USD,
         HUF = (HUF*EUR)/USD,
         LTL = (LTL*EUR)/USD,
         LVL = (LVL*EUR)/USD,
         MTL = (MTL*EUR)/USD,
         PLN = (PLN*EUR)/USD,
         RON = (RON*EUR)/USD,
         ROL = (ROL*EUR)/USD,
         SEK = (SEK*EUR)/USD,
         SIT = (SIT*EUR)/USD,
         SKK = (SKK*EUR)/USD,
         CHF = (CHF*EUR)/USD,
         ISK = (ISK*EUR)/USD,
         NOK = (NOK*EUR)/USD,
         HRK = (HRK*EUR)/USD,
         RUB = (RUB*EUR)/USD,
         TRY = (TRY*EUR)/USD,
         AUD = (AUD*EUR)/USD,
         BRL = (BRL*EUR)/USD,
         CAD = (CAD*EUR)/USD,
         CNY = (CNY*EUR)/USD,
         HKD = (HKD*EUR)/USD,
         IDR = (IDR*EUR)/USD,
         ILS = (ILS*EUR)/USD,
         INR = (INR*EUR)/USD,
         KRW = (KRW*EUR)/USD,
         MXN = (MXN*EUR)/USD,
         MYR = (MYR*EUR)/USD,
         NZD = (NZD*EUR)/USD,
         PHP = (PHP*EUR)/USD,
         SGD = (SGD*EUR)/USD,
         THB = (THB*EUR)/USD,
         ZAR = (ZAR*EUR)/USD)

Forex_DF_asc <- Forex_DF %>% arrange(ymd(Forex_DF$Date))
Forex_No_Date_DF <- Forex_DF %>% dplyr::select(-Date)

Forex_Lead_DF <- Forex_DF %>%
  mutate(EUR.lead = lead(EUR)) %>%
  mutate(JPY.lead = lead(JPY)) %>%
  mutate(BGN.lead = lead(BGN)) %>%
  mutate(CYP.lead = lead(CYP)) %>%
  mutate(CZK.lead = lead(CZK)) %>%
  mutate(DKK.lead = lead(DKK)) %>%
  mutate(EEK.lead = lead(EEK)) %>%
  mutate(GBP.lead = lead(GBP)) %>%
  mutate(HUF.lead = lead(HUF)) %>%
  mutate(LTL.lead = lead(LTL)) %>%
  mutate(LVL.lead = lead(LVL)) %>%
  mutate(MTL.lead = lead(MTL)) %>%
  mutate(PLN.lead = lead(PLN)) %>%
  mutate(RON.lead = lead(RON)) %>%
  mutate(ROL.lead = lead(ROL)) %>%
  mutate(SEK.lead = lead(SEK)) %>%
  mutate(SIT.lead = lead(SIT)) %>%
  mutate(SKK.lead = lead(SKK)) %>%
  mutate(CHF.lead = lead(CHF)) %>%
  mutate(ISK.lead = lead(ISK)) %>%
  mutate(NOK.lead = lead(NOK)) %>%
  mutate(HRK.lead = lead(HRK)) %>%
  mutate(RUB.lead = lead(RUB)) %>%
  mutate(TRY.lead = lead(TRY)) %>%
  mutate(AUD.lead = lead(AUD)) %>%
  mutate(BRL.lead = lead(BRL)) %>%
  mutate(CAD.lead = lead(CAD)) %>%
  mutate(CNY.lead = lead(CNY)) %>%
  mutate(HKD.lead = lead(HKD)) %>%
  mutate(IDR.lead = lead(IDR)) %>%
  mutate(ILS.lead = lead(ILS)) %>%
  mutate(INR.lead = lead(INR)) %>%
  mutate(KRW.lead = lead(KRW)) %>%
  mutate(MXN.lead = lead(MXN)) %>%
  mutate(MYR.lead = lead(MYR)) %>%
  mutate(NZD.lead = lead(NZD)) %>%
  mutate(PHP.lead = lead(PHP)) %>%
  mutate(SGD.lead = lead(SGD)) %>%
  mutate(THB.lead = lead(THB)) %>%
  mutate(ZAR.lead = lead(ZAR))

Forex_Lead_DF <- Forex_Lead_DF[-nrow(Forex_Lead_DF),]

Forex_Lead2_DF <- Forex_Lead_DF %>%
  mutate(EUR.lead2 = lead(EUR.lead)) %>%
  mutate(JPY.lead2 = lead(JPY.lead)) %>%
  mutate(BGN.lead2 = lead(BGN.lead)) %>%
  mutate(CYP.lead2 = lead(CYP.lead)) %>%  
  mutate(CZK.lead2 = lead(CZK.lead)) %>%
  mutate(DKK.lead2 = lead(DKK.lead)) %>%
  mutate(EEK.lead2 = lead(EEK.lead)) %>%
  mutate(GBP.lead2 = lead(GBP.lead)) %>%
  mutate(HUF.lead2 = lead(HUF.lead)) %>%
  mutate(LTL.lead2 = lead(LTL.lead)) %>%
  mutate(LVL.lead2 = lead(LVL.lead)) %>%
  mutate(MTL.lead2 = lead(MTL.lead)) %>%
  mutate(PLN.lead2 = lead(PLN.lead)) %>%
  mutate(RON.lead2 = lead(RON.lead)) %>%
  mutate(ROL.lead2 = lead(ROL.lead)) %>%
  mutate(SEK.lead2 = lead(SEK.lead)) %>%
  mutate(SIT.lead2 = lead(SIT.lead)) %>%
  mutate(SKK.lead2 = lead(SKK.lead)) %>%
  mutate(CHF.lead2 = lead(CHF.lead)) %>%
  mutate(ISK.lead2 = lead(ISK.lead)) %>%
  mutate(NOK.lead2 = lead(NOK.lead)) %>%
  mutate(HRK.lead2 = lead(HRK.lead)) %>%
  mutate(RUB.lead2 = lead(RUB.lead)) %>%
  mutate(TRY.lead2 = lead(TRY.lead)) %>%
  mutate(AUD.lead2 = lead(AUD.lead)) %>%
  mutate(BRL.lead2 = lead(BRL.lead)) %>%
  mutate(CAD.lead2 = lead(CAD.lead)) %>%
  mutate(CNY.lead2 = lead(CNY.lead)) %>%
  mutate(HKD.lead2 = lead(HKD.lead)) %>%
  mutate(IDR.lead2 = lead(IDR.lead)) %>%
  mutate(ILS.lead2 = lead(ILS.lead)) %>%
  mutate(INR.lead2 = lead(INR.lead)) %>%
  mutate(KRW.lead2 = lead(KRW.lead)) %>%
  mutate(MXN.lead2 = lead(MXN.lead)) %>%
  mutate(MYR.lead2 = lead(MYR.lead)) %>%
  mutate(NZD.lead2 = lead(NZD.lead)) %>%
  mutate(PHP.lead2 = lead(PHP.lead)) %>%
  mutate(SGD.lead2 = lead(SGD.lead)) %>%
  mutate(THB.lead2 = lead(THB.lead)) %>%
  mutate(ZAR.lead2 = lead(ZAR.lead))

Forex_Lead2_DF <- Forex_Lead2_DF[-nrow(Forex_Lead2_DF),]

# Filtered out all other European, Canadian, Australian, and New Zealand currencies,
Forex_EUR_Leads_DF <- dplyr::select(Forex_Lead2_DF,ends_with(c("Date","EUR",".lead",".lead2"))) %>% dplyr::select(!contains(c("EUR.lead","EUR.lead2","BGN","CYP", "DKK","EEK","CZK","SEK","CHF","HUF","GBP","LTL","LVL","MTL","PLN","RON","ROL","ISK","NOK","HRK","CAD","AUD","NZD","SIT","SKK","RUB")))

test_index <- createDataPartition(Forex_EUR_Leads_DF$EUR ,p=0.1, list=FALSE)
train_forex <- Forex_EUR_Leads_DF[-test_index,]
test_forex <- Forex_EUR_Leads_DF[test_index,]
```


## Data Visualization
```{r}
eur_plot <- ggplot(Forex_DF_asc,mapping=aes(x=Date,y=EUR,group=1)) +
  geom_line() +
  labs(x="Year",y="EUR/USD")
dkk_plot <- ggplot(Forex_DF_asc,mapping=aes(x=Date,y=DKK,group=1)) +
  geom_line() +
  labs(x="Year",y="DKK/USD") #Trend is nearly identical to EUR
cor(Forex_DF$EUR,Forex_DF$DKK)
aud_plot <- ggplot(Forex_DF_asc,mapping=aes(x=Date,y=AUD,group=1)) +
  geom_line() +
  labs(x="Year",y="AUD/USD") #Trend is nearly identical to EUR
cor(Forex_DF$EUR,Forex_DF$AUD)
jpy_plot <- ggplot(Forex_DF_asc,mapping=aes(x=Date,y=JPY,group=1)) +
  geom_line() +
  labs(x="Year",y="JPY/USD")
cor(Forex_DF$EUR,Forex_DF$JPY)
cny_plot <- ggplot(Forex_DF_asc,mapping=aes(x=Date,y=CNY,group=1)) +
  geom_line() +
  labs(x="Year",y="CNY/USD")
cor(Forex_DF$EUR,Forex_DF$CNY)
krw_plot <- ggplot(Forex_DF_asc,mapping=aes(x=Date,y=KRW,group=1)) +
  geom_line() +
  labs(x="Year",y="KRW/USD")
cor(Forex_DF$EUR,Forex_DF$KRW)

grid.arrange(eur_plot,dkk_plot,aud_plot)
grid.arrange(eur_plot,jpy_plot,cny_plot,krw_plot)
```
## Forecasting Setup
```{r}
# Extraneous Code from working with Forecasting Models
#
# Future_Forex_DF <- insertRows(Forex_DF_asc,1,new = NA)
# Future_Forex_DF <- insertRows(Future_Forex_DF,nrow(Future_Forex_DF),new = NA)
# 
# Future_Forex_DF <- replicate(2,{(insertRows(Future_Forex_DF,nrow(Future_Forex_DF)+1,new = NA))})
# 
# #Future_Forex_DF$Date[1] <- Sys.Date()+1
# #Forex_Lag <- lag(Forex_DF,k=1)
# #Forex_1999_2019_DF <- Forex_DF[Forex_DF$Date < "2020-01-01",]
# #Forex_2020_2022_DF <- Forex_DF[Forex_DF$Date >= "2020-01-01",]
# 
# Forex_skip_30 <- head(Forex_DF_asc,-30)
# 
# eur_arima <- auto.arima(Forex_skip_30$EUR, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
# eur_pred<- forecast(eur_arima,h=30)
# eur_pred$upper[,2]
# tail(Forex_DF_asc,30)
# postResample(eur_pred$upper[,2],tail(Forex_DF_asc$EUR,30))
```


## Subset Selection to see best variables for predicting Euro exchange rate
```{r}
eur_vars <- regsubsets(EUR~.-Date, data = Forex_EUR_Leads_DF,nvmax=ncol(Forex_EUR_Leads_DF)-1,method="backward")
summary(eur_vars)

eur_summary <- summary(eur_vars)

# plot(eur_summary$adjr2, xlab = "Number of Variables", ylab = "ADJR2", type = "l")
# plot(eur_summary$rss, xlab = "Number of Variables", ylab = "RSS", type = "l")
# plot(eur_summary$rsq, xlab = "Number of Variables", ylab = "RSQ", type = "l")
# plot(eur_summary$cp, xlab = "Number of Variables", ylab = "CP", type = "l")
# plot(eur_summary$bic, xlab = "Number of Variables", ylab = "BIC", type = "l")


which.max(eur_summary$adjr2)
which.min(eur_summary$rss)
which.max(eur_summary$rsq)
which.min(eur_summary$cp)
which.min(eur_summary$bic)

var_names <- colnames(eur_summary$which)
var_in_best <- var_names[eur_summary$which[which.min(eur_summary$bic),]]
formula <- paste0("EUR~",paste(var_in_best[-1],collapse = "+"))
var_in_best
```

## Summarizing Linear Models to see statistical significance of each variable
```{r}
summary(lm(formula,Forex_EUR_Leads_DF))
```
## Linear Model using optimal BIC Subset variables
```{r}
eur_lm <- lm(formula,train_forex)
eur_lm_pred <- predict(eur_lm,test_forex)
postResample(eur_lm_pred,test_forex$EUR)

ggplot()+
  geom_line(test_forex,mapping=aes(x=Date,y=EUR,color="Test Set",group=1)) +
  geom_line(mapping=aes(x=test_forex$Date,y=eur_lm_pred,color="Predictions",group=1)) +
  scale_color_manual("",breaks=c("Test Set","Predictions"),values=c("Blue","Red")) +
  labs(title="Linear Regression Predicition")


Forex_post_2005 <- Forex_EUR_Leads_DF %>% filter(Date>"2005-04-05")

test_2005_index <- createDataPartition(Forex_post_2005$EUR ,p=0.1, list=FALSE)
train_2005_forex <- Forex_post_2005[-test_2005_index,]
test_2005_forex <- Forex_post_2005[test_2005_index,]

eur_2005_lm <- lm(formula,train_2005_forex)
eur_2005_lm_pred <- predict(eur_2005_lm,test_2005_forex)
postResample(eur_2005_lm_pred,test_2005_forex$EUR)

ggplot()+
  geom_line(test_2005_forex,mapping=aes(x=Date,y=EUR,color="Test Set",group=1)) +
  geom_line(mapping=aes(x=test_2005_forex$Date,y=eur_2005_lm_pred,color="Predictions",group=1)) +
  scale_color_manual("",breaks=c("Test Set","Predictions"),values=c("Blue","Red")) +
  labs(title="Linear Regression (2005-Present) Predicition")


```
## PCR Model
```{r}
eur_pcr <- pcr(EUR~., data = train_forex, scale = TRUE, validation = "CV")
eur_pcr_pred <- predict(eur_pcr,test_forex,ncomp=which.min(eur_summary$bic))
postResample(eur_pcr_pred,test_forex$EUR)

ggplot()+
  geom_line(test_forex,mapping=aes(x=Date,y=EUR,color="Test Set",group=1)) +
  geom_line(mapping=aes(x=test_forex$Date,y=eur_pcr_pred,color="Predictions",group=1)) +
  scale_color_manual("",breaks=c("Test Set","Predictions"),values=c("Blue","Red")) +  
  labs(title="Principal Component Regression Predicition")

```
## Bagged Trees
```{r}
bag.mod <- bagging(
  formula =EUR~.-Date,
  data = train_forex,
  nbagg = 150,
  coob = TRUE,
  control = rpart.control(minsplit=20, cp = 0)
)

bag.mod

eur_bag_pred <- predict(bag.mod, test_forex)
postResample(eur_bag_pred,test_forex$EUR)

ggplot()+
  geom_line(test_forex,mapping=aes(x=Date,y=EUR,color="Test Set",group=1)) +
  geom_line(mapping=aes(x=test_forex$Date,y=eur_bag_pred,color="Predictions",group=1)) +
  scale_color_manual("",breaks=c("Test Set","Predictions"),values=c("Blue","Red")) +
  labs(title="Bagged Trees Predicition")

```
## Boosted Trees
```{r}
eur_boost<-gbm(EUR~.-Date, data=train_forex, 
           distribution="gaussian",
           n.trees=200, 
           interaction.depth=4, 
           shrinkage= 0.1,
           cv.folds = 10)

summary(eur_boost)
best.iter <- gbm.perf(eur_boost, plot.it = TRUE)

eur_boost_pred <- predict(eur_boost, newdata = test_forex, n.trees = best.iter)
postResample(eur_boost_pred,test_forex$EUR)

ggplot()+
  geom_line(test_forex,mapping=aes(x=Date,y=EUR,color="Test Set",group=1)) +
  geom_line(mapping=aes(x=test_forex$Date,y=eur_boost_pred,color="Predictions",group=1)) +
  scale_color_manual("",breaks=c("Test Set","Predictions"),values=c("Blue","Red")) +
  labs(title="Boosted Trees Predicition")

```

## Applying the Models
```{r}

##Bagged Trees Predictions

Apr_11_22 <- Forex_EUR_Leads_DF %>% filter(Date=="2022-04-11")
bag_pred <- predict(bag.mod,Apr_11_22)
postResample(bag_pred,Apr_11_22$EUR)

Apr_21_22 <- Forex_EUR_Leads_DF %>% filter(Date=="2022-04-21")
bag_pred <- predict(bag.mod,Apr_21_22)
postResample(bag_pred,Apr_21_22$EUR)

Apr_27_22 <- Forex_EUR_Leads_DF %>% filter(Date=="2022-04-27")
bag_pred <- predict(bag.mod,Apr_27_22)
postResample(bag_pred,Apr_27_22$EUR)

today <- Forex_EUR_Leads_DF %>% filter(Date==Sys.Date())
bag_pred <- predict(bag.mod,today)
postResample(bag_pred,today$EUR)

##Boosted Trees Predictions

Apr_11_22 <- Forex_EUR_Leads_DF %>% filter(Date=="2022-04-11")
boost_pred <- predict(eur_boost,Apr_11_22)
postResample(boost_pred,Apr_11_22$EUR)

Apr_21_22 <- Forex_EUR_Leads_DF %>% filter(Date=="2022-04-21")
boost_pred <- predict(eur_boost,Apr_21_22)
postResample(boost_pred,Apr_21_22$EUR)

Apr_27_22 <- Forex_EUR_Leads_DF %>% filter(Date=="2022-04-27")
boost_pred <- predict(eur_boost,Apr_27_22)
postResample(boost_pred,Apr_27_22$EUR)

today <- Forex_EUR_Leads_DF %>% filter(Date==Sys.Date())
boost_pred <- predict(eur_boost,today)
postResample(boost_pred,today$EUR)

Jan_6_23 <- Forex_EUR_Leads_DF %>% filter(Date=="2023-01-06")
boost_pred <- predict(eur_boost,Jan_6_23)
postResample(boost_pred,Jan_6_23$EUR)
```
